---
title: "coMethTF: functional annotation of DMRs and DMS in epigenome-wide association studies"
author:
  - Tiago Chedraoui Silva^[University of Miami Miller School of Medicine, txs902@miami.edu]
  - Lily wang^[University of Miami Miller School of Medicine, lily.wangg@gmail.com]
output: 
  prettydoc::html_pretty:
    theme: leonids
    toc: true
    df_print: paged
    code_download: false
    toc_depth: 3
bibliography: bibliography.bib    
editor_options:
  chunk_output_type: inline    
vignette: >
    %\VignetteIndexEntry{How to use MethTF}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---
  
<style> body {text-align: justify} </style>
  
```{r, include = FALSE}
knitr::opts_chunk$set(
collapse = TRUE,
comment = "#>"
)
library(dplyr)
library(sesameData)
sesameDataCacheAll()
```

# 1. Introduction

Transcription factors (TFs) are proteins that facilitate the transcription of DNA into RNA. A number of recent studies have observed that the binding of TFs onto DNA can be affected by DNA methylation, and in turn, DNA methylation can also be added or removed by proteins associated with transcription factors [@bonder2017disease; @banovich2014methylation; @zhu2016transcription].

To provide functional annotations for differentially methylated regions (DMRs) and differentially methylated CpG sites (DMS),
`coMethTF` performs integrative analyses using matched DNA methylation and gene expression along with Transcription Factor Binding Sites (TFBS) data. coMethTF evaluates, prioritizes and annotates DNA methylation regions (or sites) with high regulatory potential that works synergistically with TFs to regulate target gene expressions, without any additional ChIP-seq data. 

The results from `coMethTF` can be used to generate testable hypothesis on the synergistic interaction of DMRs and TFs in gene regulation. 
`coMethTF` can be used either to evaluate regulatory potentials of candidate regions or to search for methylation coupled TF regulatory processes in the entire genome.

# 2. Installation 

You can install the development version of the package from GitHub using the following code:

```{r, eval = FALSE}
devtools::install_github("TransBioInfoLab/coMethTF")
```

```{r, eval = TRUE, include = FALSE}
library(coMethTF)
```

After the package is successfully installed you can load its functions with:

```{r setup, eval = FALSE}
library(coMethTF)
```

# 3. Dataset

The input of `coMethTF` are DNA methylation and gene expression datasets measured on the same set of samples. Using additional information from the HOCOMOCO database [@kulakovskiy2016hocomoco], we will then create a data frame with information on the triplet: DMR (or CpG), TF and target gene. 

For illustration, we will use chromosome 21 data from 38 TCGA-COAD samples.

## 3.1 DNA methylation dataset

The DNA methylation dataset is a matrix with methylation beta or M-values. If there are potential confounding factors (e.g. batch effect, age, sex) in the dataset, this matrix can also contain residuals extracted from fitting linear regression instead (see **Section 5** below). 
The samples are in the columns and regions/probes are in the rows.   

### 3.1.1 Individual CpGs data

```{R warning=FALSE}
dna.met.chr21 <- get(data("dna.met.chr21"))
```

```{R}
dna.met.chr21[1:5,1:5]
```

```{R}
dna.met.chr21 <- map_probes_to_regions(dna.met.chr21)
```


```{R}
dna.met.chr21[1:5,1:5]
```

### 3.1.2 DMRs data
Differentially Methylated Regions (DMRs) associated with phenotypes such as tumor stage can be obtained from R packages such as \code{coMethDMR}, \code{comb-p}, \code{DMRcate} and many others. The methylation levels in multiple CpGs within the DMRs need to be summarized (e.g. using medians or Principal component analysis), then the analysis for DMR will proceed in the same way as those for CpGs, which we will illustrate next.  


## 3.2 Gene expression dataset

The gene expression dataset is a matrix with log2 transformed and normalized gene expression values. If there are potential confounding factors (e.g. batch effect, age, sex) in the dataset, this matrix can also contain residuals from linear regression instead (see **Section 5** below).
The samples are in the columns and the genes are in the rows. 

```{R}
data("gene.exp.chr21", package = "coMethTF")
gene.exp.chr21[1:5,1:5]
```

## 3.3 Creating triplet dataset
### Mapping DMR region (or CpG) to the target gene

The function `get_region_target_gene` provides two methods to link a DMR to 
a target gene, by (1) mapping the region to the closest gene (`method = "closest.gene"`) or 
(2) mapping the region to all the genes within a window (default size = 500Kbp around the region, i.e. +- 250K bp from start or end of the region) (`method = "window"`).
 
```{R}
# Get CpG information
HM450.hg38.manifest <- sesameData::sesameDataGet("HM450.hg38.manifest")
probes.gr <- make_granges_from_names(rownames(dna.met.chr21))
class(probes.gr)

# Map probes to nearest gene
mapping.closest <- get_region_target_gene(
    regions.gr = probes.gr, 
    genome = "hg38", 
    method = "closest.gene"
)
dim(mapping.closest)
mapping.closest

# Map probes to genes within 500kb window
mapping.window <- get_region_target_gene(
    regions.gr = probes.gr, 
    genome = "hg38", 
    method = "window"
)

dim(mapping.window)
mapping.window
```

Some of the columns of the data frame output from `get_region_target_gene` 
when the method is set to `"window"` are described below:

* regionID: chromossome, start, end
* regionID.extended: same as regionID, but start and end values of the regions will be 
regionID start - window.width/2 and  regionID end + window.width/2 
* window.extended.width:  width of window around region used to find target genes
* Distance.region.gene: distance from region to target gene TSS

### Mapping region to TF

The function `get_tf_in_region` maps the regions to TF using a precomputed dataset of 
HOCOMOCO motifs [@kulakovskiy2016hocomoco]. For each motif we scanned 500 bp around each EPIC and 
HM450 arrays probes. Then for each motif found in a region we extract TF within the same Tf family or subfamily,
using TFClass [@wingender2013tfclass], since they have similar binding motifs.

Therefore, a region is linked to a TF if any of the probes within it is mapped to the TF. In other words, we link the DNA methylation regions to TFs within 500 bp from the start or end of the region. 

```{R, results = "hide"}
# we will only map the first 20 regions, since this function may take a while to compute
region.tf <- get_tf_in_region(
    region = probes.gr[1:20], 
    genome = "hg38", 
    arrayType = "450k", 
    classification = "subfamily"
)
# only keep those within our example dataset
region.tf <- region.tf %>% dplyr::filter(TF_ensembl_gene_id %in% rownames(gene.exp.chr21))
```

```{R}
head(region.tf)
```


### Triplet data frame – matching DMR (or CpGs) with TFs and target genes

The triplet is a data frame with the following columns:

* `target`: gene identifier (obtained from row names of the gene expression matrix),
* `regionID`: region/cpg identifier (obtained from row names of the DNA methylation matrix)
* `TF`: gene identifier (obtained from the row names of the gene expression matrix)

```{R}
# Merging the results from the previous sections (region and target, and region and TF)
triplet <- dplyr::inner_join(
    mapping.closest[,c("regionID","ensembl_gene_id")],
    region.tf[,c("regionID","TF_ensembl_gene_id")]
    )
colnames(triplet) <- c("regionID","target","TF")
head(triplet)
```

Note that this dataset include only regions with both nearby target genes (500K bp) and nearby TFs (500 bp). 
There may also be multiple rows for a region, when multiple target gene and/or TFs are found close to it. 

# 4. Evaluating the regulatory potential of DMRs (or CpGs)

Because TF binding to DNA can be influenced by (or influence) DNA methylation levels nearby (Yin et al. 2013), target gene expression can often be influenced by the synergistic effects of both TF and DNA methylation. In other words, TF activities in gene regulation is affected by DNA methylation or vice versa. 

We will perform analyses using the 3 datasets described above in Section 3:

* DNA methylation matrix
* Gene expression matrix
* The triplet data frame with DNA methylation region, TF, and target gene

## 4.1 Analysis using model with methylation by TF interaction

The function `interaction_model` assess the regulatory impact of DNA methylation on TF regulation of target genes via two approaches: 

**Model 1** (considering DNAm values as a continuous variable) - we fit a model with TF by DNA methylation interaction using all samples.

$$log_2(RNA target)  \sim log_2(TF) + DNAm + log_2(TF) * DNAm$$
**Model 2** (considering DNAm values as a binary variable) - we define a binary group for DNA methylation values (high = 1, low = 0). That is, samples with the highest DNAm levels (top 25 percent) has high = 1, samples with lowest DNAm levels (bottom 25 pecent) has high = 0. Note that in this implementation, only samples wih DNAm values in the first and last quartiles are considered.

$$log_2(RNA target) \sim log_2(TF) + \text{DNAm Quartile Group} + log_2(TF) * \text{DNAm Quartile Group}$$
 
```{R interaction_model, message = FALSE, results = "hide"}
results <- interaction_model(
    triplet = triplet, 
    dnam = dna.met.chr21,
    exp = gene.exp.chr21
)
```

The output of `interaction_model` function will be a data frame with the following variables:

* `pval_<name>`: p-value for a tested variable (methylation or TF), given the other variables included in the model.
* `estimate_<name>`: estimated effect for a variable. If estimate > 0, increasing values of the variable corresponds to increased outcome values (target gene expression). If estimate < 0, increasing values of the variable correspond to decreased target gene expression levels.


The following columns are provided for the results of fitting **Model 1** to triplet data:

* `pval_met`: p-value for DNA methylation variable
* `pval_rna.tf` : p-value for TF expression
* `pval_met.rna.tf`: : p-value for DNA methylation by  TF interaction
* `estimate_met`: estimate for DNA methylation variable
* `estimate_rna.tf`: estimate  for TF expression
* `estimate_met.rna.tf`: estimate for DNA methylation by  TF interaction

```{R interaction_model_results}
dim(results)
colnames(results)

# Results for full model (with all samples)
results %>% dplyr::select(
    c(1,4,5,grep("quant",colnames(results),invert = TRUE))
    ) %>% head
```


The following columns are provided for the results of fitting **Model 2** (quartile model) to triplet data:

* `quant_pval_metGrp`: p-value for DNA methylation variable
* `quant_pval_rna.tf` : p-value for TF expression
* `quant_pval_metGrp:rna.tf`: : p-value for DNA methylation by  TF interaction
* `quant_estimates_metGrp`: estimate for DNA methylation variable
* `quant_estimates_rna.tf`: estimate  for TF expression
* `quant_estimates_metGrp:rna.tf`: estimate for DNA methylation by  TF interaction


```{R}
# Results for quartile model
results %>% dplyr::select(
  c(1,4,5,grep("quant",colnames(results)))
  ) %>% head
```

To select significant DNAm by TF intereaction from both models, 
```{r, include=FALSE, eval=FALSE}
results <- results[results$`pval_met:rna.tf`< 0.05 & results$`quant_pval_metGrp:rna.tf`< 0.05 ,]
```

## 4.2 Visualization of data  

The function `plot_interaction_model` will create figures to visualize the data, 
in a way that corresponds to the linear model we considered above. 
It requires the output from the function `interaction_model` (a dataframe), 
the DNA methylation matrix and the gene expression matrix as input. 


```{R plot_interaction_model, message = FALSE, results = "hide"}
plots <- plot_interaction_model(
    triplet.results = results[1,], 
    dnam = dna.met.chr21, 
    exp = gene.exp.chr21
    )
```

```{R, fig.height = 14, fig.width = 12}
plots[[1]]
```

The first row of the figures shows pairwise associations between DNA methylation, 
TF and target gene expression levels. 

The second row of the figures show how much TF activity on target 
gene expression levels vary by DNA methylation levels. When TF by methylation 
interaction is significant (Section 4.1), we expect the association between TF 
and target gene expression vary depending on whether DNA methylation is low or high. 

In this example, when DNA methylation is low, more abundant TF corresponds to 
decreased gene expression (a repressor TF). On the other hand, when DNA methylation 
level is high, target gene expression is relatively independent of the amount of TF available. 
One possibility is that DNA methylation might affect TF binding in this case. 


If you want to highlight different group of samples, you can also use the metadata argument
with a dataframe with rownames as samples (same order as the columns of the DNAm and gene expression),
and a column highlighting the groups.


```{R, message = TRUE, results = "hide"}
data("clinical")
metadata <- clinical[,c("sample_type"),drop = FALSE]
```

```{R, eval = TRUE}
plots <- plot_interaction_model(
        triplet.results = results[1,,drop = FALSE], 
        dnam = dna.met.chr21, 
        exp = gene.exp.chr21,
        metadata = metadata
)
```


```{R, fig.height = 14, fig.width = 12, eval = TRUE}
plots[[1]]
```

# 5. Controlling effects from confounding variables

Both gene expressions and DNA methylation levels can be affected by age, sex, shifting in cell types, batch effects and other confounding variables.  In this section, we illustrate analysis workflow that reduces confounding effects, by first extracting the residual data with the function 
`get_residuals`, before fitting the models discussed above in Section 4. 

The `get_residuals` function will use gene expression (or DNA methylation data) and phenotype data as input. 
To remove confounding effects in gene expression data, we use the `get_residuals` function which extract residuals after fitting the following model: 
`features ~ covariate_{1} + covariate_{2} ... + covariate_{N}`

```{R residuals, results = "hide", eval = TRUE}
data("gene.exp.chr21")
data("clinical")
metadata <- clinical[,c("sample_type","gender")]

gene.exp.chr21.residuals <- get_residuals(gene.exp.chr21[1:250,], metadata)
```

```{R, eval = TRUE}
gene.exp.chr21.residuals[1:5,1:5]
```

```{R, results = "hide", eval = TRUE}
data("dna.met.chr21")
dna.met.chr21 <- map_probes_to_regions(dna.met.chr21)
dna.met.chr21.residuals <- get_residuals(dna.met.chr21[1:5,],metadata)
```

```{R, eval = TRUE}
dna.met.chr21.residuals[1:5,1:5]
```

The `interaction_model` function can then be applied to these residuals data, as the following:  
`rna.target ~ met + rna.tf + rna.tf * met`.


```{R, message = FALSE, results = "hide", eval = TRUE}
results <- interaction_model(
    triplet = triplet, 
    dnam = dna.met.chr21.residuals, 
    exp = gene.exp.chr21.residuals
)
```

# 5. Session information
```{R,size = 'tiny'}
sessionInfo()
```

# 6. Bibliography
